---
title: "Introduction to R for Data Science Part III"
author: "Justin Post"
date: "August 12-13, 2019"
output: 
      ioslides_presentation:
         css: css/style.css
         widescreen: true
transition: faster
logo: img/logo.png
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
options(dplyr.print_min = 5)
library(tidyverse)
library(haven)
rm(iris)
```

## What do we want to be able to do?

- Read in data 

- **Manipulate data**  

- Plot data  

- Summarize data  

- Analyze data  
<!--Don't want to get lost in the details.  What are our goals? what is the point of this?-->


## Schedule    

**Day 1**  

- Install R/R studio  
- R Studio Interface
- Classes and Objects    
- Attributes and Basic Data Object Manipulation    
- Reading in Data/Writing Out Data
- **Logical Statements and Subsetting/Manipulating Data**

<!--For the rest of today and probably the start of tomorrow we will look at logical statements and subsetting/manipulating data.  Very useful!-->


## Logical Statements

- Logical statement - comparison of two quantities  
  + resolves as `TRUE` or `FALSE`  

<div style="float: left; width: 45%;">
```{r}
#Use of ==, !=, >=, <=, >, <
"hi" == " hi"
"hi" == "hi"
4 == 1
```
</div>
<div style="float: right; width: 45%;">
```{r}
4 >= 3
4 != 1
"hi" != "hello"
```
</div>


## Logical Statements

- Logical statement - comparison of two quantities  
  + resolves as `TRUE` or `FALSE`  

<div style="float: left; width: 45%;">
```{r}
#use of is. functions
is.numeric("Word")
is.numeric(10)
```
</div>
<div style="float: right; width: 45%;">
```{r}
is.character("10")
is.na(c(1:2, NA, 3))
```
</div>

## Logical Statements/Subsetting Data

- Useful for indexing a vector
```{r}
iris <- tbl_df(iris)
iris
```


## Logical Statements/Subsetting Data

- Useful for indexing a vector
- Standard way for subsetting data (or use `subset` function)
```{r,eval=FALSE}
iris[iris$Species == "setosa", ]
```

- Concept:
    + Feed index a vector of `TRUE`/`FALSE` or `0`/`1` values  
    + R returns elements where `TRUE` or `1` occurred


## Logical Statements/Subsetting Data

- Useful for indexing a vector
```{r}
##Obtain a vector that indicates which rows are "setosa" species
iris$Species == "setosa"
```

## Logical Statements/Subsetting Data

- Useful for indexing a vector
- Standard way for subsetting data (or use `subset` function)
```{r}
iris[iris$Species == "setosa", ]
```


## Logical Statements/Subsetting Data

- Useful for indexing a vector
- Standard way for subsetting data (or use `subset` function)
```{r}
subset(iris, Species == "setosa")
```



## Logical Statements/Subsetting Data

- Useful for indexing a vector
- Optional way: `filter()` from `dplyr` (installed with `tidyverse`)

```{r}
filter(iris, Species == "setosa")
```


## Logical Statements/Subsetting Data

```{r}
filter(iris, Species != "setosa")
```
- We'll spend a good bit of time with dplyr in a bit!

## Implicit Data Change  
**Aside: Coercion**

- R attempts to coerce data into usable form when necessary  
- Ex: Atomic vector - all elements must be the same type
```{r}
#coerce numeric to string
c("hi", 10)
#coerce TRUE/FALSE to numeric
c(TRUE, FALSE) + 0
```


## Implicit Data Change  
**Aside: Coercion**

- R attempts to coerce data into usable form when necessary  
- Coerce from less flexible to more flexible
    + Data types from least to most flexible:
        * logical
        * integer
        * double
        * character.
```{r}
#logical to character
c(TRUE, "hi")
```

## Implicit Data Change  
**Aside: Coercion**

- R attempts to coerce data into usable form when necessary  
- Explicit coercion with `as.` functions

<div style="float: left; width: 48%;">
```{r}
as.numeric(c(TRUE, FALSE, TRUE))
mean(c(TRUE, FALSE, TRUE))
```
</div>
<div style="float: right; width: 48%;">
```{r}
as.character(c(1, 2, 3.5, TRUE))
```
</div>

<div style="float: right; width: 50%;">
- Why does TRUE return "1"?
</div>


## Logical Statements
**Logical Operators**

- `&` 'and'
- `|` 'or'

Operator | A,B true       | A true, B false   | A,B false
---------|----------------|-------------------|--------------
`&`      | `A & B = TRUE` | `A & B = FALSE`   | `A & B = FALSE`
`|`      | `A | B = TRUE` | `A | B = TRUE`    | `A | B = FALSE`


## Logical Statements
**Logical Operators**

- `&` 'and'
- `|` 'or'

Operator | A,B true       | A true, B false   | A,B false
---------|----------------|-------------------|--------------
`&`      | `A & B = TRUE` | `A & B = FALSE`   | `A & B = FALSE`
`|`      | `A | B = TRUE` | `A | B = TRUE`    | `A | B = FALSE`

- `&&` and `||` are alternatives  
  + Looks at only first comparison if given a vector of comparisons
  

## Logical Statements
**Logical Operators**

```{r}
set.seed(3)
x <- runif(n = 10, min = 0, max = 1); x
(x < 0.25) | (x > 0.75)
(x < 0.25) || (x > 0.75)
```

## Subsetting Data

- Only pull out large petal setosa flowers
```{r}
filter(iris, (Petal.Length > 1.5) & (Petal.Width > 0.3) & 
         (Species == "setosa"))
```

## Subsetting Data

What's the idea for the filter function?

> - Condition evaluates a vector of `TRUE`/`FALSE`   

> - Returns values where TRUE is present


## Subsetting Data

What's the idea for the filter function?

 - Condition evaluates a vector of `TRUE`/`FALSE`  
 
 - Returns values where TRUE is present


```{r}
(iris$Petal.Length > 1.5) & (iris$Petal.Width > 0.3) &
  (iris$Species == "setosa")
```


## Subsetting Data

- Only pull out large petal setosa flowers
```{r}
filter(iris, (Petal.Length > 1.5) & (Petal.Width > 0.3) & 
         (Species == "setosa"))
```
- In `[, ]` notation this is more work!
```{r, eval = FALSE}
iris[(iris$Petal.Length > 1.5) & (iris$Petal.Width > 0.3) & 
         (iris$Species == "setosa"), ]
```


## Conditional Execution  
**If then, If then else**

- Often want to execute statements conditionally (say to create a new variable)  
- `if` `then` else concept
```{r,eval=FALSE}
if (condition) {
  then execute code
} 

#if then else
if (condition) {
  execute this code  
} else {
  execute this code
}
```

## Conditional Execution  
**If then, If then else**

- Often want to execute statements conditionally (say to create a new variable)  
- `if` `then` else concept
```{r,eval=FALSE}
#Or more if statements
if (condition) {
  execute this code  
} else if (condition2) {
  execute this code
} else if (condition3) {
  execute this code
} else {
  #if no conditions met
  execute this code
}
```


## Conditional Execution  
**If then, If then else**

<div style="float: left; width: 45%;">
```{r}
#silly example
a <- 5
if (a < 10){
  print("hi")
}
```
</div>

<div style="float: right; width: 45%;">
```{r}
if (a < 10){
  print("hi")
} else if (a < 40){
  print("goodbye")
} else {
  print("aloha")
}
```
</div>


## Conditional Execution  
**If then, If then else**

```{r}
#silly example
a <- 20
if (a < 10){
  print("hi")
}

if (a < 10){
  print("hi")
} else if (a < 40){
  print("goodbye")
} else {
  print("aloha")
}
```


## Conditional Execution  
**If then, If then else**

```{r}
#silly example
a <- "string"
if (a < 10){
  print("hi")
}

if (a < 10){
  print("hi")
} else if (a < 40){
  print("goodbye")
} else {
  print("aloha")
}
```


## Manipulating Data

- Logical statments great for data filtering    

> - Quite useful for creating new variables too    

> - Issue: `if` condition can only take in a single comparison  

> - Create new variable for **Large Setosa** flowers


## Manipulating Data

- Logical statments great for data filtering    

 - Quite useful for creating new variables too    

 - Issue: `if` condition can only take in a single comparison  

 - Create new variable for **Large Setosa** flowers

```{r,error=TRUE}
if ((iris$Petal.Length > 1.5) & (iris$Petal.Width > 0.3) &
      (iris$Species == "setosa")) {
   "Large Setosa"
}
```

## Manipulating Data

- Logical statments great for data filtering    

- Quite useful for creating new variables too    

- `ifelse()` is **vectorized** if statement (see help)  

- Returns a vector

```{r,eval=FALSE}
#syntax
ifelse(vector_condition, if_true_do_this, if_false_do_this)
```

## Manipulating Data

- Create new variable for **Large Setosa** flowers
```{r,error=TRUE}
ifelse((iris$Petal.Length > 1.5) & (iris$Petal.Width > 0.3) & 
         (iris$Species == "setosa"), "L-S", "NotL-S")
```


## Manipulating Data


- `dplry` has a nice function called `transmute()`  
```{r}
transmute(iris, Size = ifelse(
	(Petal.Length > 1.5) & (Petal.Width > 0.3) & (Species == "setosa"), "LS", "NotLS")
	)
```

## Manipulating Data

- `dplry` function `mutate()` does the same but *adds variable to existing data frame*
```{r}
mutate(iris, Size = ifelse(
	(Petal.Length > 1.5) & (Petal.Width > 0.3) & (Species == "setosa"), "LS", "NotLS")
	)
```


## Recap!
**Logical Operators**

- `&` 'and'  

- `|` 'or'  

- `if (condition) { ... }`  

- `if (condition) { ... } else if (condition) { ... }`  

- `ifelse(vector_condition,result_if_true,result_if_false)`  


## Subsetting/Manipulating Data
**Overview of dplyr package**

- `dplry` package made for most standard data manipulation tasks    

- Part of `tidyverse`  

- Make sure `library(tidyverse)` has been run  
 
 
## `dplyr` package

- Basic commands 
    + `tbl_df()` - convert data frame to one with better printing
    + `filter()` - subset **rows**
    + `arrange()` - reorder **rows**
    + `select()` - subset **columns**
    + `mutate()` - add newly created **column**
    + `transmute()` - create new variable
    + `group_by()` - group **rows** by a variable  
    + `summarise()` - apply basic function to data
    + `left_join()`, `right_join()`, `inner_join()`, `full_join()` - commands to combine multiple data frames


## Tidyverse Syntax

- Reason to prefer `dplyr` and packages from the `tidyverse` 

> - Fast!

> - Good defaults

> - All packages have similar syntax!  All work on `tibbles` (data frames)

> - Syntax: function(data.frame, actions, ...)


## Subsetting/Manipulating Data
`tbl_df()` - convert data frame to one with better printing

- If data read in with haven, readxl, or readr probably in this format!
- Just 'wrap' data frame

```{r,eval=FALSE}
#install.packages("Lahman")
library(Lahman)
head(Batting, n = 4) #look at just first 4 observations
```

## Subsetting/Manipulating Data
`tbl_df()` - convert data frame to one with better printing

```{r,eval=TRUE,echo=FALSE}
library(Lahman)
```
```{r}
head(Batting, n = 4) #look at just first 4 observations
```

## Subsetting/Manipulating Data

```{r}
Batting <- tbl_df(Batting)
Batting
```

## Subsetting/Manipulating Data
`filter()` - subset **rows**

- Use `filter()` to obtain only PIT data
```{r}
filter(Batting, teamID == "PIT")
```

## Subsetting/Manipulating Data
`filter()` - subset **rows**

- Multiple filters
```{r}
filter(Batting, teamID == "PIT" & yearID == 2000)
```

## Subsetting/Manipulating Data
`arrange()` - reorder **rows**

```{r}
#reorder by teamID
arrange(Batting, teamID)
```

## Subsetting/Manipulating Data
`arrange()` - reorder **rows**

```{r}
#get secondary arrangement as well
arrange(Batting, teamID, G)
```

## Subsetting/Manipulating Data
`arrange()` - reorder **rows**

```{r}
#descending instead
arrange(Batting, teamID, desc(G))
```


## Subsetting/Manipulating Data
**Piping or Chaining**

- Applying multiple functions: nesting hard to parse!
- Piping or Chaining with `%>%` operator helps

```{r}
arrange(filter(Batting, teamID == "PIT"), desc(G))
```

## Subsetting/Manipulating Data
**Piping or Chaining**

- Applying multiple functions: nesting hard to parse!
- Piping or Chaining with `%>%` operator helps

```{r}
Batting %>% filter(teamID == "PIT") %>% arrange(desc(G))
```

## Subsetting/Manipulating Data
**Piping or Chaining**

- Applying multiple functions: nesting hard to parse!
- Piping or Chaining with `%>%` operator helps
- If `dplyr` or `magrittr` package loaded, can use with other functions
```{r}
a<-runif(n = 10)
a
```

## Subsetting/Manipulating Data
**Piping or Chaining**

```{r}
#silly example
a %>% quantile()
a %>% quantile() %>% range()
```


## Subsetting/Manipulating Data
`select()` - subset **columns**

- Often only want select variables (saw `$` and `[ , ]`)
- `select()` function has same syntax as other `dplyr` functions!

```{r}
#Choose a single column by name
Batting %>% select(X2B)
```

## Subsetting/Manipulating Data
`select()` - subset **columns**

- Many ways to select variables 

```{r}
#all columns between
Batting %>% select(X2B:HR)
```

## Subsetting/Manipulating Data
`select()` - subset **columns**

- Many ways to select variables 

```{r}
#all columns containing
Batting %>% select(contains("X"))
```


## Subsetting/Manipulating Data
`select()` - subset **columns**

- Many ways to select variables 

```{r}
#all columns starting with
Batting %>% select(starts_with("X"))
```


## Subsetting/Manipulating Data
`select()` - subset **columns**

- Many ways to select variables 

```{r}
#all columns ending with
Batting %>% select(ends_with("ID"))
```

## Subsetting/Manipulating Data
`mutate()` - add newly created **column**  

`transmute()` - create new variable

```{r}
##Create an Extra Base Hits variable
Batting %>% mutate(ExtraBaseHits = X2B + X3B + HR)
```

## Subsetting/Manipulating Data
`mutate()` - add newly created **column**  

`transmute()` - create new variable

```{r}
#can't see it!
Batting %>% mutate(ExtraBaseHits = X2B + X3B + HR) %>% select(ExtraBaseHits)
```


## Subsetting/Manipulating Data
`mutate()` - add newly created **column**  

`transmute()` - create new variable

```{r}
#transmute will keep the new variable only
Batting %>% transmute(ExtraBaseHits = X2B + X3B + HR)
```

## Subsetting/Manipulating Data

- Basic data summarizations often done by groups  

- Average score by Age group  

- Median income by Education level  

- Number of participants for each Race  


## Subsetting/Manipulating Data
`group_by()` - group **rows** by a variable  

`summarise()` - apply basic function to data

- Summarization - find avg \# of doubles (X2B) 
- Remove `NA`'s
- `NA` = Not Available (R's missing data indicator)
```{r}
#average # of doubles for all players in data set
Batting %>% summarise(AvgX2B = mean(X2B, na.rm = TRUE))
```

## Subsetting/Manipulating Data
`group_by()` - group **rows** by a variable  

`summarise()` - apply basic function to data

- Summarization - find avg \# of doubles (X2B) *by team*  (explore this idea more later)
```{r}
Batting %>% group_by(teamID) %>% summarise(AvgX2B = mean(X2B, na.rm = TRUE))
```

## Subsetting/Manipulating Data  

- May want to combine two data sets: `left_join()`, `right_join()`, `inner_join()`, `full_join()`   

(Cite: http://rpubs.com/justmarkham/dplyr-tutorial-part-2)

```{r}
# create two simple data frames
a <- data_frame(color = c("green", "yellow", "red"), num = 1:3)
b <- data_frame(color = c("green", "yellow", "pink"), size = c("S", "M", "L"))
```

<div style="float: left; width: 45%;">
```{r}
a
```
</div>
<div style="float: right; width: 45%;">
```{r}
b
```
</div>

## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- Only include observations found in both "a" and "b" (automatically joins on variables that appear in both tables)

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
inner_join(a, b)
```
</div>


## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- include observations found in either "a" or "b"  

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
full_join(a, b)
```
</div>


## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- include all observations found in "a", match with b

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
left_join(a, b)
```
</div>


## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- include all observations found in "b", match with a

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
right_join(a, b)
```
</div>

## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- `right_join(a, b)` is identical to `left_join(b, a)` except for column ordering

<div style="float: left; width: 45%;">
```{r}
right_join(a,b)
```
</div>
<div style="float: right; width: 45%;">
```{r}
left_join(b, a)
```
</div>


## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- filter "a" to only show observations that match "b"

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
semi_join(a, b)
```
</div>


## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- filter "a" to only show observations that don't match "b"

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 32%;">
```{r}
anti_join(a, b)
```
</div>

## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- sometimes matching variables don't have identical names

<div style="float: left; width: 50%;">
```{r}
b <- b %>% rename(col = color)
a
```
</div>
<div style="float: right; width: 45%;">
```{r}
b
```

## Subsetting/Manipulating Data
`left_join()`, `right_join()`, `inner_join()`, `full_join()` - combine multiple DFs

- specify that the join should occur by matching "color" in "a" with "col" in "b"

<div style="float: left; width: 28%;">
```{r}
a
```
</div>
<div style="float: left; width: 28%;">
```{r}
b
```
</div>
<div style="float: right; width: 36%;">
```{r}
inner_join(a, b, 
	by = c("color" = "col"))
```
</div>


## Overview of `dplyr` package [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

- Basic commands 
    + `tbl_df()` - convert data frame to one with better printing
    + `filter()` - subset **rows**
    + `arrange()` - reorder **rows**
    + `select()` - subset **columns**
    + `mutate()` - add newly created **column**
    + `transmute()` - create new variable
    + `group_by()` - group **rows** by a variable  
    + `summarise()` - apply basic function to data
    + `left_join()`, `right_join()`, `inner_join()`, `full_join()` - commands to combine multiple data frames
    

## Activity 
- [**Manipulating Data Activity** instructions](http://www4.stat.ncsu.edu/~post/DataScienceR/ManipulateActivity.html) available on web  

- Work in small groups  

- Ask questions!  TAs and I will float about the room  

- Feel free to ask questions about anything you didn't understand as well!   

