---
title: "Exercises 10 - Bar Plots Solutions"
output:
  github_document
---

## 10

Recall the Lahman package that has data about baseball players. Consider the "Master" data set. Write R code to do the following:

1. Convert the Master data set to a tibble
```{r}
library(Lahman)
myMaster <- as_tibble(Master)
```

2. Create a table that gives the frequency of players by birth months.
```{r}
table(myMaster$birthMonth)
```

3. Create a table that gives the proportion of players by birth months. (Proportion = # in category/total #)
```{r}
tab <- table(myMaster$birthMonth)
tab <- tab / sum(tab)
tab
```

4. Create a bar chart of the player birth month frequencies. Give the plot appropriate labels.
```{r}
g <- ggplot(myMaster %>% drop_na(birthMonth), aes(x = as.factor(birthMonth)))
g + geom_bar() + 
	labs(x = "Birth Month", title = "Bar Graph of Player Birth Month") +
	scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

5. Create a two-way contingency table that gives the frequencies of how the player bats ("bats") and the way the player throws ("throws").

```{r}
twoWayTable <- table(myMaster$bats, myMaster$throws)
twoWayTable
```

6. Create a stacked bar chart of the player bat and throw preferences. Give the plot appropriate labels.

```{r}
g <- ggplot(myMaster %>% drop_na(bats, throws), aes(x = bats))
g + geom_bar(aes(fill = throws)) + labs(title = "Bar Graph of Handedness and Batting Style")
```

<hr>

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
