---
title: "Exercises 3 - Common Data Objects Solutions"
output:
  github_document:
    df_print: paged
---

## 3.1 Vectors

1) Create a vector `vec` with elements 2, 4, 6, ..., 12. 

```{r}
vec <- seq(from = 2, to = 12, by = 2)
#or
vec <- (1:6)*2
```

2) Write code to see the 1st, 3rd, and 5th elements of `vec` in the console.

```{r}
vec[c(1, 3, 5)]
```

3) You can exclude elements from a vector by using a '`-`' sign in front of their indices. Utilize the `length()` function to see the `vec` object in the console without its last element.

```{r}
vec[-length(vec)]
```

## 3.2 Subsetting data frames

1) The `iris` dataframe is a built-in R object.  Write code to show the 4th column of the `iris` dataset in the console (there are many ways to do this!).

```{r}
## using column index
iris[, 4]

## using column name
iris[, "Petal.Width"]

## using column name and $ operator
iris$Petal.Width

## like a list
iris[[4]]
```

2) Write code to see just the first five rows and just the `Species` and `Sepal.Width` columns of the `iris` object in the console.

```{r}
iris[1:5, c("Sepal.Width", "Species")]
```

3) Write code to see `iris` dataset in the console with the columns displayed in alphabetical order.  This can be done using the `sort()` function.  Use the result of `sort(colnames(iris))` along with `[]` to select the columns in alphabetical order.

```{r}
iris[, sort(colnames(iris))]
```

<hr>

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
