---
title: "Exercises 2 - Objects Solutions"
output:
  github_document:
    df_print: paged
---

## 2.1 Creating R Objects

1. The code below creates a vector of three character strings (more on vectors shortly).  Use this code and the 'storage arrow' to create an object names `adj`.  

`c("scary","intelligent","new")`

Solution: 
```{r}
adj <- c("scary","intelligent","new")
adj
```

2. Similarly, create an object called `nouns` using the storage arrow and code below.  

`c("bugs", "beings", "houses")`

Solution: 
```{r}
nouns <- c("bugs", "beings", "houses")
nouns
```

3. The `paste()` function is useful for combining character strings.  After completing the above parts, run the following:

```{r}
paste(adj, nouns)
```

<hr> 

## 2.2 Investigating Objects

1. `iris` is a built-in R object (that means you have it in your environment even if it doesn't show it).  Determine the `class` and structure (`str`) of the `iris` object.

Solution: 
```{r}
class(iris)
str(iris)
```

2.  Run the code below to create a quick visual of this dataset. Note that R is determining what to do with the `plot()` function based on the object given to it!

```{r}
plot(iris)
```

<hr>

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
