---
title: "Exercises 6 - dplyr Solutions"
output:
  github_document
---

We'll continue to work on the same .Rmd file from the previous exercise.    

<hr>  

# 6.1 Fix the BabyNameSupp Data

Edit your code chunk that reads the data in from a csv file and add the argument `col_types = cols(sex = col_character())` to the `read_csv()` code.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(dcData)

data("BabyNames", package = "dcData")
```

```{r}
BabyNamesSupp <- 
  read_csv("https://github.com/jbpost2/Basics-of-R-for-Data-Science-and-Statistics/raw/master/datasets/BabyNamesSupp.csv",
           col_types = cols(sex = col_character()))  # fixes `sex`
```


# 6.2 Combining Data Sets with `dplyr::bind_rows()`

1. Write a quick note below your last R code chunk that mentions that we are going to combine the two data sets into one.  

2. Use the `bind_rows()` function from `dplyr` to combine `BabyNames` & `BabyNamesSupp` via `bind_rows(BabyNames, BabyNamesSupp)` into an object called `BabyNamesFull`.

```{r}
BabyNamesFull <- bind_rows(BabyNames, BabyNamesSupp)
```

<hr> 

# 6.3 Filter and Arrange Rows

- Filter the `BabyNamesFull` data to include only the names we are interested in (Hint: use `%in%`).  Just have this print out to the console.   

- Continue the previous code to then reorder the rows descending on the `count` variable.  

```{r}
# vector of names
names <- c("Justin", "Alexander", "Jacob", "Anderson", "George")

BabyNamesFull %>%
  filter(name %in% names) %>% 
  head() #This is just added so we don't have too much print out

BabyNamesFull %>%
  filter(name %in% names) %>% 
  arrange(desc(count)) %>%
  head() #Added for less printing
```

<hr> 

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
