---
title: "Exercises 11 - ggplot"
output:
  github_document
---

We'll continue to work on the same .Rmd file from the previous exercise.    

<hr> 

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(dcData)

data("BabyNames", package = "dcData")
BabyNamesSupp <- 
  read_csv("https://github.com/jbpost2/Basics-of-R-for-Data-Science-and-Statistics/raw/master/datasets/BabyNamesSupp.csv",
           col_types = cols(sex = col_character()))  # fixes `sex`
BabyNamesFull <- bind_rows(BabyNames, BabyNamesSupp)
```


1. Filter the data by the names of interest we have, group the data by name, and find the total counts using the code below:


```{r}
names <- c("Justin", "George", "Alexander", "Jacob", "Anderson")
barData <- BabyNamesFull %>%
  filter(name %in% names) %>% 
  group_by(name) %>%
  summarise(total = sum(count))
```



2. Use the data object created to create a bar plot with names on the x-axis and the total count on the y-axis.  You'll need to specify both the x and y variables  and change the statistic to "identity" in the `geom_bar()` function.  

```{r}
ggplot(barData) +
  geom_bar(stat = "identity", aes(x = name, y = total)) 
```

3.  Use the data object created by the code below to create a filled bar plot with the sex variable denoting the coloring of columns.

```{r}
barDataSex <- BabyNamesFull %>%
  filter(name %in% c("Shannon", "Lindsay", "Avery")) %>% 
  group_by(name, sex) %>%
  summarise(total = sum(count))
```

```{r}
ggplot(barDataSex) +
 geom_bar(stat = "identity", aes(x = name, y = total, fill = sex)) 
```
<hr>

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
