---
title: "Exercises 4 - Reading Data Solutions"
output:
  github_document
---

## 4.1 Read data contained in an R package.

1) Load the `dcData` and `tidyverse` packages. Use the 
  - Note: `dcData` is installed from GitHub, so it requires an extra step.        - `devtools::install_github("mdbeckman/dcData")`
  - this might take a while (30+ sec?) depending on your Internet connection
  After loading in the `dcData` package, read the `BabyNames` data from `dcData` into your R environment using the `data()` function as follows:  
  `data(BabyNames)`  
Now you should see the `BabyNames` object in the Environment pane  
    - Double click the object name to open a spreadsheet View  
    - Spreadsheet view includes basic sorting capability!  

```{r}
# devtools::install_github("mdbeckman/dcData")

library(tidyverse)  # this actually loads a group of packages all at once
library(dcData)     
data(BabyNames)
#or
data("BabyNames", package = "dcData")
```

<hr> 

## 4.2 Read a .csv file from the web

The file "BabyNameSupp.csv" includes a few years of more recent data to augment the `BabyNames` data. The file can be downloaded straight to R from the link here:  

https://github.com/jbpost2/Basics-of-R-for-Data-Science-and-Statistics/raw/master/datasets/BabyNamesSupp.csv

Read this file in using `read_csv()` from the `readr` package.  Save the data as an object called `BabyNamesSupp`.

Note: Reading the data **will** produce a warning message!  Read the warning message carefully; what seems to have gone wrong?

- Open the spreadsheet view to investigate `BabyNamesSupp`  
- Can you confirm what seems to have gone wrong during the data intake?

```{r}
BabyNamesSupp <- read_csv("https://github.com/jbpost2/Basics-of-R-for-Data-Science-and-Statistics/raw/master/datasets/BabyNamesSupp.csv")
```

<hr>

<br>

<a href = "https://jbpost2.github.io/Basics-of-R-for-Data-Science-and-Statistics/CourseFiles.html"><button type="button">Course Files Page</button></a>
