---
title: "Exercises to Accompany TeachingWithR Workshop"
subtitle: "2021 US Conference on Teaching Statistics"
author: "Matthew Beckman & Justin Post"
date: "June 25, 2021"
output:
  github_document:
    df_print: paged
---


# Part 2. Import Data

<hr>

## 2.1 Packages

- **Task 1:** call the `dcData` and `tidyverse` packages   
    - Note: `dcData` is installed from GitHub, so it requires an extra step.  You may have already done this from the instructions prior to the workshop, but it is shown again here if needed:
    - `devtools::install_github("mdbeckman/dcData")`
    - this might take a while (30+ sec?) depending on your Internet connection

<hr>

### Solution

```{r}
## from instructions before workshop, if needed
# devtools::install_github("mdbeckman/dcData")

library(tidyverse)  # this actually loads a group of packages all at once
library(dcData)     
```

<hr>

## 2.2 Data Intake: `BabyNames`

- **Task 1:** Read `BabyNames` data from `dcData` into your R environment using the `data()` function  
    - Locate the `BabyNames` object in the Environment pane  
    - Double click the object name to open a spreadsheet View  
    - Spreadsheet view includes basic sorting capability  

Use the spreadsheet view to answer the following:  

- **Task 2:** how many total rows of data are included?  

- **Task 3:** describe what each row in the data *actually* represents.

- **Task 4:** find the largest `count` in the `BabyNames` data.  How would you interpret this result?

- **Task 5:** what are the max & min available `year` in the data?

<hr>

### Solution 


Task 2.2.1: 

```{r}
data("BabyNames", package = "dcData")
```


Task 2.2.2: about 1.8 million rows.  We can see this from environment tab, spreadsheet view (footer), among other places

Task 2.2.3: The frequency of a given name, within a year, associated with a `sex`

Task 2.2.4: According to US Social Security records, 1947 was a banner year to name a female Linda!

Task 2.2.5: `BabyNames` data spans 1880 through 2013

<hr>

## 2.3 Data Intake: `BabyNamesSupp` 

The file "BabyNameSupp.csv" includes a few years of more recent data to augment the `BabyNames` data. Run the starter code shown below to read the data and complete the tasks.  

**Important:** The starter code **will** produce a warning message!  Don't worry, it's part of the exercise!


```{r}
# starter code for BabyNamesSupp 
BabyNamesSupp <- read_csv("https://jbpost2.github.io/TeachingWithR/datasets/BabyNamesSupp.csv")
```


- **Task 1:** Read the warning message carefully; what seems to have gone wrong?

- **Task 2:** open the spreadsheet view to investigate `BabyNamesSupp`... 
    - Which years of data are available to us in `BabyNamesSupp`?
    - Can you confirm what seems to have gone wrong during the data intake?

- **Task 3:** use the following R functions to investigate `BabyNamesSupp` further:
    - `head(BabyNamesSupp)`
    - `tail(BabyNamesSupp)`
    - `str(BabyNamesSupp)`


- **Task 4 (Challenge):** Why did `read_csv( )` seem to have this problem with the data intake? Any ideas how we might fix it?
    - Hint: "RStudio >> Help >> read_csv" 
    - don't worry about *actually* fixing it yet, just try to identify root cause.

At this point, we aren't attempting to prepare the `BabyNamesSupp` data for analysis.  We're just reading it into the R environment and making observations.  We'll be using these data again in later exercises, so we will make the necessary corrections at that point.

<hr>

### Solution

Task 2.3.1: The warning indicates a "parsing failure" and shows that R was expecting logical elements (e.g., TRUE/FALSE) and found "M" in many cases

Task 2.3.2: Yup, R interpreted "F" (meaning female in our data) to mean FALSE and then could not handle "M" appropriately as a result.  These data include years 2014 through 2019.

Task 2.3.3: 

```{r}
head(BabyNamesSupp)
tail(BabyNamesSupp)
str(BabyNamesSupp)
```


Task 2.3.4 (Challenge): The help documentation explains that `read_csv( )` guesses each variable type based on the first 1000 records.  The data source had apparently sorted results such that the first 1000 (or more) rows were all "F" so `read_csv( )` concluded these to be logical (e.g., TRUE or FALSE) data.  Quite a reasonable default under most circumstances, but shows why we should always carefully inspect our data intake! 

<hr>

## 2.4 Help Documentation for Data Sets

Search "RStudio >> Help" to learn about the data...

- **Task 1:** what can you learn about `BabyNames` data from RStudio Help?
- **Task 2:** what can you learn about `BabyNamesSupp` data from RStudio Help?  What happened?

<hr>

### Solution 

Task 2.4.1: Lots! Sice this is loaded from an R package, the package author can make additional information available to you that describes the data, explains the variables, provides a source, etc.

Task 2.4.2: Nothing!  The data were read from an external file, so R doesn't have any information at all to share with us.  

<hr>

## 2.5 Data for 2020 (Challenge)

**Task 1:** Want to include 2020 data too?  See if you can locate it, read the data into R, and review the data intake (hint: `BabyNames` help documentation includes a source to investigate).

Again, we aren't attempting to process the 2020 data yet.  We're just reading it into the R environment and making observations about that process.  We'll be using this data again later in the exercises, so we will make the necessary corrections at that point.

<hr>

## Solution

Task 2.5.1 (Challenge): visit "RStudio >> Help >> BabyNames >> Source" and then download the zip file to find the 2020 data inside.  This can be read into R using `read_csv( )` even though the file extension is `.txt` since the data are comma delimited.  You may note that the structure of the data differs slightly from our two other sources, so we will need some new tools to process it for use.

```{r}

## challenge
# locate zip file including all years of data, identify 2020 and read into R:
BabyNames2020 <- 
    read_csv("https://jbpost2.github.io/TeachingWithR/datasets/yob2020.txt", 
             col_names = FALSE)


```

<hr>

# Part 3. R Markdown 

Note: you might hang onto the RStudio default text provided in the new R Markdown file for the moment... it's packed with tiny examples that will come in handy!

[coming up next...]

<hr>


<br>

<a href = "https://jbpost2.github.io/TeachingWithR/CourseFiles.html"><button type="button">Course Files Page</button></a> 

